---
title: "Clustering"
author: "Laura.w and Kelsey Gonzalez"
date: "12/18/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r global_options, echo=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE,fig.width=5)
```


Load packages
```{r}
library(tidyverse)    # data manipulation and visualization
library(kernlab)      # SVM methodology
library(e1071)        # SVM methodology
library(ISLR)         # contains example data set "Khan"
library(RColorBrewer) # customized coloring of plots
library(performanceEstimation) # for performance estimation of svm
library(RCurl) # for performance estimation of svm
library(factoextra) #for hopkins statistic
library(clustertend) #for hopkins statistic
library(cluster) # for gower similarity, pam, and diana (a divisive hierarchical method). clara() is also included, which is basically pam by sampling
library(Rtsne) # for t-SNE plot
library(ggrepel)
library(dbscan)
library(rpart)
library(rpart.plot)
library(DMwR2)
library(rsample)
library(adabag)
library(ipred)
library(randomForest)
library(gbm)
library(performanceEstimation)
library(e1071)
library(nnet)
options(scipen = 999)

```

#Data Preprocessing

Review of World Health Organization's Life Expectancy Data taken from [Kaggle](https://www.kaggle.com/kumarajarshi/life-expectancy-who).

```{r}
rm(list=ls())
load("LifeExpectancyData.Rdata")
data$Status = as.factor(data$Status)
data = subset(data,select = -c(`Hepatitis B`))
```

Then we aquire data to include continent values
```{r}
# from https://github.com/lukes/ISO-3166-Countries-with-Regional-Codes
ISO3166 <- read.csv(text = (getURL("https://raw.githubusercontent.com/lukes/ISO-3166-Countries-with-Regional-Codes/master/all/all.csv")))
ISO3166$name <- as.character(ISO3166$name)
# They use Eswatini instead of Swaziland, so let's update that & other naming mismatches
ISO3166$name[ISO3166$name == "Moldova, Republic of"] <- "Republic of Moldova"
ISO3166$name[ISO3166$name == "Korea, Republic of"] <- "Republic of Korea"
ISO3166$name[ISO3166$name == "Korea (Democratic People's Republic of)"] <- "Democratic People's Republic of Korea"
ISO3166$name[ISO3166$name == "Congo, Democratic Republic of the"] <- "Democratic Republic of the Congo"
ISO3166$name[ISO3166$name == "North Macedonia"] <- "The former Yugoslav republic of Macedonia"
ISO3166$name[ISO3166$name == "Tanzania, United Republic of"] <- "United Republic of Tanzania"
```

We created a country data frame to make the data join more easily, and then joined them by country("country", "name"), while adding the continent ("region")
```{r}
countries <- as.data.frame(unique(data$Country))
colnames(countries) <- "Country" 
countries$Country <- as.character(countries$Country)

dfcountry <- left_join(x = countries, 
                   y = ISO3166, 
                   by = c("Country" = "name"))


data <- left_join(x = data,
                   y = dfcountry[,c("Country","region")], 
                   by = c("Country" = "Country"))
```

```{r}
data$Status = as.factor(data$Status)
data$Country = as.factor(data$Country)

data <- data %>% select(-c(Polio, "percentage expenditure", Measles, `HIV/AIDS`)) 


dataavg <- data %>% 
  drop_na %>% 
  group_by(Country) %>% 
  summarize(Status = last(Status),
            Life = mean(`Life expectancy`),
            AdultMort = mean(`Adult Mortality`),
            infantdeath = mean(`infant deaths`),
            Alcohol = mean(Alcohol),
            BMI = mean(BMI),
            under_five_death = mean(`under-five deaths`),
            total_exp = mean(`Total expenditure`),
            Diptheria = mean(Diphtheria),
            GDP = mean(GDP),
            Pop = mean(Population),
            thinness_1_19  = mean(`thinness  1-19 years`),
            Income_comp = mean(`Income composition of resources`),
            Schooling = mean(Schooling),
            region = last(region)) 
summary(dataavg)

```

```{r, message=FALSE}

set.seed(5)

ten_countries <- dataavg %>% 
  select(Country) %>% 
  sample_n(.,14) 
```

```{r, echo=FALSE, message = FALSE}
dataavg %>% 
  mutate(label = ifelse(Country %in% ten_countries$Country, 
                        as.character(Country), 
                        "")) %>% 
  ggplot(aes(BMI, Life)) +
  geom_point(size = 3.5, 
             alpha = .9,
             shape = 21, 
             col = "white", 
             fill = "#0162B2") + 
  geom_text_repel(aes(label = label),
                  size = 4.5,
                  point.padding = .2,
                  box.padding = .3,
                  force = 1,
                  min.segment.length = 0) +
  theme_minimal(14) +
  theme(legend.position = "none",
        panel.grid.minor = element_blank()) +
  labs(x = "BMI",
       y = "life expectancy") +
  geom_smooth()
```

```{r, echo=FALSE, message = FALSE}
dataavg %>% 
  mutate(label = ifelse(Country %in% ten_countries$Country, 
                        as.character(Country), 
                        "")) %>% 
  ggplot(aes(log_GDP, Life)) +
  geom_point(size = 3.5, 
             alpha = .9,
             shape = 21, 
             col = "white", 
             fill = "#0162B2") + 
  geom_text_repel(aes(label = label),
                  size = 4.5,
                  point.padding = .2,
                  box.padding = .3,
                  force = 1,
                  min.segment.length = 0) +
  theme_minimal(14) +
  theme(legend.position = "none",
        panel.grid.minor = element_blank()) +
  labs(x = "logGDP",
       y = "life expectancy")+
  geom_smooth()
```

```{r, echo=FALSE, message = FALSE}
dataavg %>% 
  mutate(label = ifelse(Country %in% ten_countries$Country, 
                        as.character(Country), 
                        "")) %>% 
  ggplot(aes(log_infant_deaths, Life)) +
  geom_point(size = 3.5, 
             alpha = .9,
             shape = 21, 
             col = "white", 
             fill = "#0162B2") + 
  geom_text_repel(aes(label = label),
                  size = 4.5,
                  point.padding = .2,
                  box.padding = .3,
                  force = 1,
                  min.segment.length = 0) +
  theme_minimal(14) +
  theme(legend.position = "none",
        panel.grid.minor = element_blank()) +
  labs(x = "Infant Deaths (logged)",
       y = "life expectancy")+
  geom_smooth()
```

```{r, echo=FALSE, message = FALSE}
dataavg %>% 
  mutate(label = ifelse(Country %in% ten_countries$Country, 
                        as.character(Country), 
                        "")) %>% 
  ggplot(aes(Diptheria, Life)) +
  geom_point(size = 3.5, 
             alpha = .9,
             shape = 21, 
             col = "white", 
             fill = "#0162B2") + 
  geom_text_repel(aes(label = label),
                  size = 4.5,
                  point.padding = .2,
                  box.padding = .3,
                  force = 1,
                  min.segment.length = 0) +
  theme_minimal(14) +
  theme(legend.position = "none",
        panel.grid.minor = element_blank()) +
  labs(x = "Diptheria",
       y = "life expectancy")+
  geom_smooth()
```

```{r, echo=FALSE, message = FALSE}
dataavg %>% 
  mutate(label = ifelse(Country %in% ten_countries$Country, 
                        as.character(Country), 
                        "")) %>% 
  ggplot(aes(log_Population, Life)) +
  geom_point(size = 3.5, 
             alpha = .9,
             shape = 21, 
             col = "white", 
             fill = "#0162B2") + 
  geom_text_repel(aes(label = label),
                  size = 4.5,
                  point.padding = .2,
                  box.padding = .3,
                  force = 1,
                  min.segment.length = 0) +
  theme_minimal(14) +
  theme(legend.position = "none",
        panel.grid.minor = element_blank()) +
  labs(x = "Population (Logged)",
       y = "life expectancy")+
  geom_smooth()


```

```{r, echo=FALSE, message = FALSE}
dataavg %>% 
  mutate(label = ifelse(Country %in% ten_countries$Country, 
                        as.character(Country), 
                        "")) %>% 
  ggplot(aes(AdultMort, Life)) +
  geom_point(size = 3.5, 
             alpha = .9,
             shape = 21, 
             col = "white", 
             fill = "#0162B2") + 
  geom_text_repel(aes(label = label),
                  size = 4.5,
                  point.padding = .2,
                  box.padding = .3,
                  force = 1,
                  min.segment.length = 0) +
  theme_minimal(14) +
  theme(legend.position = "none",
        panel.grid.minor = element_blank()) +
  labs(x = "Adult Mortality",
       y = "life expectancy")+
  geom_smooth()

```

#Descriptive Analyses
## Pam Clustering 


First things first, do our data even have a natural clustering tendency?
the larger the result, the higher clustering tendency

The hopkins statistic on our dataset shows about a .79 clustering tendency, where in some cases can resemble 90% confidence level.
source: https://www.datanovia.com/en/lessons/assessing-clustering-tendency/
Compute Hopkins statistic for dataset:n must be no larger than num of samples

```{r}
# Hopkins, for numerical data only 

data_no_cat <- dataavg %>% 
  select(-Country, -Status, -region) %>% 
  as.matrix()
rownames(data_no_cat) <- dataavg$Country

res <- get_clust_tendency(data_no_cat, n=139, graph = TRUE)
res$hopkins_stat

```

```{r}
res$plot
```
Our Next try shows the clustering tendency at .25 which is very small; < 0.5, no clustering tendency
```{r}
set.seed(78)
res <- hopkins(data_no_cat, n = 137)
res$H 
```

Calculate distance between attributes using gower.
```{r}
gower_dist <- daisy(dataavg[, -1], 
                    metric = "gower")


# Check attributes to ensure the correct methods are being used (I = interval, N = nominal)
# Note that despite logratio being called, the type remains coded as "I"

# cleaning it up
gower_mat <- as.matrix(gower_dist)
rownames(gower_mat) <- colnames(gower_mat) <- dataavg$Country

```

Calculate Silhouette width for 2 to 10 clusters using PAM - The top cluster options are 2 and 3
```{r}
sil <- c(NA)
for(i in 2:10){
  pam_fit <- pam(gower_mat, diss=TRUE, k=i)
  sil[i] <-pam_fit$silinfo$avg.width
}

plot(1:10, sil,
xlab = "Number of clusters",
ylab = "Silhouette Width",
lines(1:10, sil))
```

review 2 clusters
```{r}
pam_fit <- pam(gower_dist, diss = TRUE, k = 2)

# add cluster labels to the data. We will use result1 later
data_clean <- data.frame(dataavg, pam_fit$cluster)

# show clustering results by country
result1 <- data_clean %>% select(Country,pam_fit.cluster)
clusplot(pam_fit)

```

Review 3 clusters
```{r}
pam_fit <- pam(gower_dist, diss = TRUE, k = 3)

# add cluster labels to the data. We will use result1 later
data_clean <- data.frame(dataavg, pam_fit$cluster)

# show clustering results by country
result1 <- data_clean %>% select(Country,pam_fit.cluster)
clusplot(pam_fit)
```

group_by cluster and then compute the summary data (means, median, etc) for each cluster
```{r}
data_results <- data_clean %>%
  mutate(cluster = pam_fit.cluster) %>% #add the cluster column
  select(-pam_fit.cluster) %>% 
  group_by(cluster) %>% #group countries by its cluster 
  do(the_summary = summary(.)) #do: summarize by group/cluster,add the_summary column

data_results$the_summary
```

The results suggest:

Cluster 1 has all developing countries (53) with a median Life Expectancy of around 58 years, and is mainly comprised of 38 countries in Africa, 13 in Asian, 1 in the Americas, and 1 in Oceana.

Cluster 2 has all developing countries (55) with a median Life Expectancy of around 73 years, and is mainly comprised of 22 countries in the Americas, 17 in Asia, 6 in Oceana, 5 in Africa, and 5 in Europe.

Cluster 3 has mostly developed countries (26) and 6 developing countries with a median Life Expectancy of around 80 years, and is mainly comprised of countries in Europe, 29, and Oceania, 1.




One way to visualize many variables in a lower dimensional space is with t-distributed stochastic neighborhood embedding, or t-SNE. 
```{r}
tsne_obj <- Rtsne(gower_dist, is_distance = TRUE)

tsne_data_pam <- tsne_obj$Y %>%
  data.frame() %>% 
  setNames(c("X", "Y")) %>%
  mutate(cluster = factor(pam_fit$clustering),
         name = data_clean$Country)

ggplot(aes(x = X, y = Y), data = tsne_data_pam) +
  geom_point(aes(color = cluster)) + 
  geom_text(data=subset(tsne_data_pam[pam_fit$medoids, ]), # add names of centroids in
            aes(X,Y,label=name))
```

clearly pam isn't the ideal clustering algorithm here. 

## Hierarchical Clustering 

We can use hierarchical clustering to show where the best fit clusters might be.
```{r}
h <- hclust(d = as.dist(gower_dist), method="complete")
plot(h, cex = 0.3, hang = -1, main="Hierarchal Cluster of Predicted Probabilities") +
  abline(h=quantile(gower_dist, .75), col="blue")
```

```{r}
tsne_data_hc <- tsne_obj$Y %>%
  data.frame() %>% 
  setNames(c("X", "Y")) %>%
  mutate(cluster = factor(cutree(h, h=quantile(gower_dist, .75))),
         name = data_clean$Country)

set.seed(8)
labels <- tsne_data_hc %>%
    group_by(cluster) %>%
    sample_n(.,1) 

tsne_data_hc <- tsne_data_hc %>% 
  mutate(label = ifelse(name %in% labels$name, as.character(name), ""))

ggplot(aes(x = X, y = Y), data = tsne_data_hc) +
  geom_point(aes(color = cluster)) + 
  geom_text(aes(X,Y,label=label))

```

## Density Clustering with Optics


```{r}
#16 dimensions, so set k = 17

kNNdistplot(gower_dist, k=17) +
  abline(h=0.2, col='red', lty=2)

res_col <- optics(gower_dist, eps=10, minPts = 7)

plot(res_col)
```

Density clustering with DBSCAN
```{r}
#4 clusters
res_col_d <- extractDBSCAN(res_col, eps_cl=0.117)
plot(res_col_d)
```

```{r}
tsne_data_den_d<- tsne_obj$Y %>%
  data.frame() %>% 
  setNames(c("X", "Y")) %>%
  mutate(cluster = factor(res_col_d$cluster),
         name = data_clean$Country)

set.seed(5)
labels <- tsne_data_den_d %>% 
    filter(cluster != "0") %>% 
    group_by(cluster) %>%
    sample_n(.,1) 

tsne_data_den_d <- tsne_data_den_d %>% 
  mutate(label = ifelse(name %in% labels$name, as.character(name), ""))

ggplot(aes(x = X, y = Y), data = tsne_data_den_d) +
  geom_point(aes(color = cluster)) + 
  geom_text(aes(X,Y,label=label)) +
  scale_color_manual(values=c("#999999",
                              "#F8766D",
                              "#00BCD8",
                              "#E76BF3",
                              "#6BB100",
                              "#619CFF"))
```
The values are all over the place, making sense why these are "outliers" according to optical clustering. 

30 countries are outliers, but which?

```{r}
tsne_data_den_d %>% 
    filter(cluster == "0") %>% 
    select(name) 


dataavg %>% filter(Country %in% as.vector(tsne_data_den_d$name[tsne_data_den_d$cluster == 0])) %>% summary()
# the values are all over the place, making sense why these are "outliers" according to optical clustering. 

```