---
title: 'R exercise 1: Know Your Data'
author: "Kelsey Gonzalez and Laura Werthmann"
date: "September 15, 2019"
due date: "September 18, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load the data 
```{r}
load("LifeExpectancyData.Rdata")

head(data)

```
 


# Central Tendency Measures: mean, median, and mode
### mean
Overall mean life expectancy 
```{r}
mean(data$`Life expectancy`, na.rm=TRUE)
```
Mean life expectancy by year
```{r}
aggregate(data$`Life expectancy`, list(data$Year), mean, na.rm=TRUE)

```

Mean life expectancy overtime by country
```{r}
aggregate(data$`Life expectancy`, list(data$Country), mean, na.rm=TRUE)

```

Mean life expectancy by country status
```{r}
aggregate(data$`Life expectancy`, list(data$Status), mean, na.rm=TRUE)

```


## median
```{r}

```


### mode
```{r}

# installed.packages("modes")
library("modes")
modes(data$`Life expectancy`, type=1, digits="NULL", nmore="NULL")

```



# statistics of spread

### variance
```{r}

```

### standard deviation
```{r}

```
  
### inter-quartile range
```{r}

```

# obtain summaries on subset of data


```{r}

```

```{r}

```

# data visualization: plotting in R


```{r}

```


```{r}

```


```{r}

```
