---
title: 'R exercise 1: Know Your Data'
author: "Kelsey Gonzalez and Laura Werthmann"
date: "September 15, 2019"
due date: "September 18, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


```

Load the data 
```{r}
load("LifeExpectancyData.Rdata")



head(data)

```
 


# Central Tendency Measures: mean, median, and mode
### mean
Overall mean life expectancy 
```{r}
mean(data$`Life expectancy`, na.rm=TRUE)
```
Mean life expectancy by year
```{r}
aggregate(data$`Life expectancy`, list(data$Year), mean, na.rm=TRUE)

```

Mean life expectancy overtime by country
```{r}
aggregate(data$`Life expectancy`, list(data$Country), mean, na.rm=TRUE)

```

Mean life expectancy by country status
```{r}
aggregate(data$`Life expectancy`, list(data$Status), mean, na.rm=TRUE)

```


### median
```{r}
# installed.packages("DMwR2")
library(DMwR2)
centralValue(data$`Life expectancy`)

```


### mode
```{r}

# installed.packages("modes")
library("modes")
modes(data$`Life expectancy`, type=1, digits="NULL", nmore="NULL")

```



# statistics of spread

### variance
```{r}
var(data$`Life expectancy`, na.rm=TRUE)
```

### standard deviation
```{r}
sd(data$`Life expectancy`, na.rm=TRUE)

```
  
### inter-quartile range
```{r}
range(data$`Life expectancy`, na.rm=TRUE)
max(data$`Life expectancy`, na.rm=TRUE)
min(data$`Life expectancy`, na.rm=TRUE)
IQR(data$`Life expectancy`, na.rm=TRUE)
```

```{r}
quantile(data$`Life expectancy`, na.rm=TRUE)
quantile(data$`Life expectancy`, probs=c(0.2, 0.8), na.rm=TRUE)
```

# obtain summaries on subset of data


```{r}
# install.packages("Hmisc")
library(Hmisc)
describe(data)

```

```{r}
library(dplyr)
group_by(data, Country, Status) %>% summarise(nObs=n(), mA7=median(`Life expectancy`))


```

# data visualization: plotting in R

```{r ggplot=plot, echo = FALSE}
suppressPackageStartupMessages(library(ggplot2))
ggplot(data=LifeExpectancy,
       mapping = aes(x = LifeExpectancy$`Life expectancy`, 
                     y = LifeExpectancy$'Year')) +
geom_jitter()+
  xlab ('Life expectancy')
  ylab ('Year')
```



```{r ggplot=plot, echo = FALSE}
suppressPackageStartupMessages(library(ggplot2))
ggplot(data = LifeExpectancy,
       mapping = aes(LifeExpectancy$`Life expectancy`)) +
  geom_histogram(binwidth = 5)

```


```{r ggplot=plot, echo = FALSE}
suppressPackageStartupMessages(library(ggplot2))
ggplot(data = LifeExpectancy,
       mapping = aes(group=LifeExpectancy$Status,
                     x=LifeExpectancy$`Life expectancy`,
                     y=LifeExpectancy$Year)) +
  geom_boxplot() +
  xlab ('Life Expectancy')
  ylab ('Year')
```

##{r ggplot=plot, echo=FALSE}
suppressPackageStartupMessages(library(ggplot2))
ggplot(data = LifeExpectancy,
       aes(group=LifeExpectancy$Status,
           x=LifeExpectancy$`Life expectancy`,
           y=LifeExpectancy$Year)) +
  geom_boxplot() + facet_grid(LifeExpectancy$GDP~LifeExpectancy$Status)
  
  
```{r ggplot=plot, echo=FALSE}
suppressPackageStartupMessages(library(ggplot2))
ggplot(data = LifeExpectancy,
       aes(x=LifeExpectancy$`Life expectancy`, y=LifeExpectancy$GDP)) +
  geom_point() + ggtitle("Relationship btwn Life Expectancy and GDP")
  

```


```{r}
plot(LifeExpectancy$`Life expectancy`, LifeExpectancy$Population, main="Relationship btwn Life Expectancy and Population")

```

```{r}
ggplot(data = LifeExpectancy,
       aes(x=LifeExpectancy$`Life expectancy`,
           y=LifeExpectancy$`percentage expenditure`)) +
  geom_point()+ggtitle("Relationship btwn LE and Percentage Expenditure") +facet_grid(.~LifeExpectancy$Status)
```

```{r}

```


```{r}

```


```{r}

```
